---
title: "Modeling"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Introduction

```{r}
#load packages needed
library(tidyverse)
library(ggplot2)
library(caret)
library(Metrics)
```

```{r}
#read in csv data
data_original <- read_csv('diabetes_binary_health_indicators_BRFSS2015.csv')
str(data_original)
```

Next, we can check whether there are duplicated data in this dataset.

```{r}
deplicates <- data_original[duplicated(data_original), ]
nrow(deplicates)
#from the output, we see that there are 24206 duplicates

#then we need to exclude data that are duplicated
#dataset data is updated here
data_no_duplicates <- data_original[!duplicated(data_original), ]
nrow(data_no_duplicates)
#from the output, there are total 229,474 rows in the updated dataset data
```

Now we want to convert numeric variables to factors according to previous sorted unique values for each variable.

```{r}
data <- data_no_duplicates

data$Diabetes_binary <- factor(
  data$Diabetes_binary, 
  levels = c(0,1), 
  labels = c("No diabetes", "Prediabetes or diabetes")
)

data$HighBP <- factor(
  data$HighBP,
  levels = c(0,1), 
  labels = c("No high BP", "High BP")
)

data$HighChol <- factor(
  data$HighChol,
  levels = c(0,1), 
  labels = c("No high cholesterol", "High cholesterol")
)

data$CholCheck <- factor(
  data$CholCheck,
  levels = c(0,1), 
  labels = c("No cholesterol check", "Cholesterol check")
)

data$Smoker <- factor(
  data$Smoker,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$Stroke <- factor(
  data$Stroke,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$HeartDiseaseorAttack <- factor(
  data$HeartDiseaseorAttack,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$PhysActivity <- factor(
  data$PhysActivity,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$Fruits <- factor(
  data$Fruits,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$Veggies <- factor(
  data$Veggies,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$HvyAlcoholConsump <- factor(
  data$HvyAlcoholConsump,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$AnyHealthcare <- factor(
  data$AnyHealthcare,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$NoDocbcCost <- factor(
  data$NoDocbcCost,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$GenHlth <- factor(
  data$GenHlth,
  levels = c(1, 2, 3, 4, 5), 
  labels = c("Excellent", "Very good", "Good", "Fair", "Poor")
)

data$DiffWalk <- factor(
  data$DiffWalk,
  levels = c(0,1), 
  labels = c("No", "Yes")
)

data$Sex <- factor(
  data$Sex,
  levels = c(0,1), 
  labels = c("Female", "Male")
)

data$Age <- factor(
  data$Age,
  levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13), 
  labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80 or older")
)

data$Education <- factor(
  data$Education,
  levels = c(1, 2, 3, 4, 5, 6), 
  labels = c("Never attended school or only kindergarten", "Elementary", "Some high school", "High school graduate", "Some college or technical school", "College graduate")
)

data$Income <- factor(
  data$Income,
  levels = c(1, 2, 3, 4, 5, 6, 7, 8), 
  labels = c("Less than $10,000", "$10,000 to less than $15,000", "$15,000 to less than $20,000", "$20,000 to less than $25,000", "$25,000 to less than $35,000", "$35,000 to less than $50,000", "$50,000 to less than $75,000", "$75,000 or more")
)

str(data)
```

## Split Data

Split this data into a training and test set. Before modeling, let’s scale and centralized data.

```{r}
set.seed(11)


trainIndex <- createDataPartition(data$Diabetes_binary, p = .7,
                                  list = FALSE,
                                  times = 1)

train_data <-  data[trainIndex, ]
test_data <- data[-trainIndex, ]

#check the dimensions of our training data and testing data frame
dim(train_data)
dim(test_data)

pre_proc_values <- preProcess(train_data, method = c("center", "scale"))

#Scaling and centralizing train and test data sets.
train_transformed <- predict(pre_proc_values, train_data)
test_transformed <- predict(pre_proc_values, test_data)
```

## Logistic Regression Models

### Logistic regression Model 1

```{r}
set.seed(11)

trctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 3)

logistic_M1_fit <- train(Diabetes_binary ~ ., 
                         data = train_transformed, 
                         method = "glm",
                         family="binomial",
                         trControl=trctrl)
summary(logistic_M1_fit)
```

### Logistic regression Model 2

```{r}
set.seed(11)

logistic_M2_fit <- train(Diabetes_binary ~ . - NoDocbcCost - Fruits - AnyHealthcare + PhysHlth:GenHlth + PhysHlth:DiffWalk + GenHlth:DiffWalk + Income:Education + Income:GenHlth + Income:DiffWalk,
                         data = train_transformed, 
                         method = "glm",
                         family="binomial",
                         trControl=trctrl)
summary(logistic_M2_fit)
```

### Logistic regression Model 3

```{r}
set.seed(11)

logistic_M3_fit <- train(Diabetes_binary ~ GenHlth + HighBP + DiffWalk + BMI + HighChol + Age + HeartDiseaseorAttack + PhysHlth + Income + PhysActivity, 
                         data = train_transformed, 
                         method = "glm",
                         family="binomial",
                         trControl=trctrl)
summary(logistic_M3_fit)
```

### Models comparison

Obtain log-loss value for each model:

```{r}
#using logLoss  as our metric for logistic M1 model:

#get predicted values corresponding to the probabilities that each observation in test data belongs to "Prediabetes or diabetes"
predicted_prob_M1 <- predict(logistic_M1_fit,
                           newdata = select(test_transformed, -Diabetes_binary), type='prob')

#convert variable Diabetes_binary from factor to numeric in order to use logLoss function
log_loss_logistic_M1 <- logLoss(as.numeric(as.character(test_transformed$Diabetes_binary) == "Prediabetes or diabetes"),
                              predicted_prob_M1$`Prediabetes or diabetes`)

#using logLoss  as our metric for logistic M2 model:

predicted_prob_M2 <- predict(logistic_M2_fit,
                           newdata = select(test_transformed, -Diabetes_binary), type='prob')

log_loss_logistic_M2 <- logLoss(as.numeric(as.character(test_transformed$Diabetes_binary) == "Prediabetes or diabetes"),
                              predicted_prob_M2$`Prediabetes or diabetes`)

#using logLoss  as our metric for logistic M3 model:

predicted_prob_M3 <- predict(logistic_M3_fit,
                           newdata = select(test_transformed, -Diabetes_binary), type='prob')

log_loss_logistic_M3 <- logLoss(as.numeric(as.character(test_transformed$Diabetes_binary) == "Prediabetes or diabetes"),
                              predicted_prob_M3$`Prediabetes or diabetes`)

#list of accuracy obtained from each logistic model
list(log_loss_logistic_M1 = log_loss_logistic_M1, log_loss_logistic_M2 = log_loss_logistic_M2, log_loss_logistic_M3 = log_loss_logistic_M3)
```

From the result after using logLoss(), we can see that all 3 logistic models have similar log-loss values (between 0.34 to 0.35). Therefore, considering the complexity of models, model 3 could be chosen as the best model among these three models since it is the simplest model among them.

## Classification Tree

```{r}

```

## Random Forest

## Final Model Selection
